PLAE enables quick and easy searching across one million ocular transcriptomes

### 2022-03-22 manuscript_00.Rmd
many text adds

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst("~/data/scEiaD/2021_03_17_meta_filter.fst") %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType))
# remove eric bharti 2D/3D RPE
meta_filter <- fst::read_fst("/Volumes/McGaughey_S/scEiaD_v3/2021_11_09_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType))

@@

## scEiaD re-created and pipeline updated

The scEiaD pipeline is largely the same as in Swamy et al. [@swamy_building_2021]. Briefly, we collect the raw fastq files and quantify with the kallisto with the bustools system [@melsted_barcode_2019; @melsted_modular_2019]. The quantified counts are merged into one large matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, cluster finding, UMAP finding. We use a xgboost-based machine learning approach to transfer cell type labels to unlabelled cells.

The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination. Third, we altered the procedure for automatically aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in batch correction.

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify in a combined data analysis, we must align shared genes. We use a multi-step procedure.

##

## pan mouse and pan human as non-ocular reference
Tabula Muris and

@@

## Updated the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in four important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 300 to 450, which reduces the number of cells but improves the quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

After the dataset and quality control updates the scEiaD v02 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

@@

While PLAE and Spectacle both contain large numbers of ocular-related scRNA datasets, they are structured differently. PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together. This allows for queries to be constructed across studies and

The Spectacle web app contains access 87,305 single cell transcriptomes across four publications.

@@

## Rich web functionality

## Differential testing identifies thousands of genes enriched in various ocular cell types

##

# Conclusion

>>>>

library(pool)
library(RSQLite)
meta_filter_v0 <- fst::read_fst("/Volumes/McGaughey_S/scEiaD/paper_data/2021_03_17_meta_filter.fst") %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precurs' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType))
# remove eric bharti 2D/3D RPE
meta_filter <- fst::read_fst("/Volumes/McGaughey_S/scEiaD_v3/2021_11_09_meta_filter.fst") %>% filter(study_accession != 'Bharti_Nguyen_iRPE_2D_3D') %>% 
  mutate(CellType_predict = case_when(CellType_predict == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType_predict),
         CellType= case_when(CellType == 'AC/HC_Precursor' ~ 'AC/HC Precursor',
                                      TRUE ~ CellType))

@@

We added X studies from X species and X tissues. 
## scEiaD re-created and pipeline updated

The scEiaD pipeline is largely the same as in Swamy et al. [@swamy_building_2021]. Briefly, we collect the raw fastq files and quantify with the kallisto / bustools system [@melsted_barcode_2019; @melsted_modular_2019]. The quantified counts are merged into one matrix that is used for downstream processing with scvitools to batch correct the data and Seurat for quality control, clustering, and 2D UMAP visualization. We use our previous xgboost-based machine learning approach to transfer cell type labels to unlabelled cells [@swamy_building_2021].

The following alterations were made: First, the cutoff to retain a cell was raised from 200 unique genes quantified to 300 Second, we use the DecontX algorithm [@yang_decontamination_2020] from the celda R package (version 1.9.2) to automatically remove ambient RNA contamination. Third, we altered the procedure for automatically aligning gene names across species (see below for details). Fourth, we use the scANVI (scvitools based) batch correction method, which leverages the known cell type labels in the batch correction process.

Two custom cutoffs for number of unique genes identified were used for the E12 chicken data and the cornea dataset (SRP362101) as an abnormally high number of cells were returned from our pipeline. We set the cutoffs for these studies to be 1000 and 800, respectively, as these values resulted in approximately the same number of cells being returned as the authors reported in their papers.

@@

As we have four species in scEiaD (*Gallus gallus*, *Macaca fascicularis*, *Mus musculus*, and *Homo sapiens*) that we unify, we must align shared genes. We found homologs/orthologs by pulling from BioMart database "Ensembl Genes 105" and using *Homo sapiens* as the reference. We then selected orthologous gene names from *Gallus gallus*, *Macaca fascicularis*, and *Mus Musculus*. This generates a table linked by Ensembl gene IDs and gene names. We then use a full join on non-duplicated Ensembl IDs between mouse and human. As we noticed a small number of gene names failed to be linked in this manner we ran another join on the remaining non-aligned genes, using gene name. This procedure gave us 17,769 shared human and mouse genes. To align the other species we again joined on Ensembl gene ID, removing genes where one chicken or macaque gene aligned to multiple human/mouse genes. In cases where multiple chicken or macaque genes aligned to one human/mouse gene, then we aggregated the counts by sum. In cases where there was no corresponding chicken or macaque gene, we filled in zeros. After these steps we had, as expected, 17,769 genes.

## Metadata curation

Every study brought into scEiaD was hand curated to label biological sample, published paper ID (where available), organ, tissue (e.g. Cornea), source (iPSC based or tissue), single cell platform (e.g. 10Xv1 or DropSeq). We also, where possible, annotated retina region (e.g. fovea), sex, and age or developmental stage. To bring across individual cell type labels (e.g. Rod, Muller Glia), we wrote custom code for each study that made this table available to link to the individual cells. We curated two types of cell type labels: "CellType" and "PubCellType" where the former is our normalized cell types taken from the published labels. In this field we normalized names, for example changin "MG" to "Muller Glia" and also dropping more detailed cell type assignments like off and on bipolar cells for Bipolar Cells (as few studies went into this level). We did retain the original published (except for the broader name normalization) labels under PubCellType.

## Pan mouse and human non-ocular references

The Tabula Muris and

## scEiaD structure and web app optimization

The web app at plae.nei.nih.gov can display million plus cell level gene expression in only a few seconds. This speed is only possible due to custom data structures to optimize for data querying. Like eyeIntegration's EiaD database, scEiaD is a sqlite database with tables for the core categories. This allows for app start up to complete in under 30 seconds on a modestly performant cloud server with memory usage under 8 GB. For the gene by cell expression matrix, the data was transformed into a "long" format where there are three columns: Gene, Cell Barcode, and Gene Count Value. This allows for cell - gene level queries to be completed, again with minimal memory usage, in under 0.5 seconds. To generate the aggregated information (in the Dot Plot, Expression Plot, and tables) we pre-calculated all queries by running a dplyr "group by" operation on all column fields. On user given query, the data is further aggregated to level request. This allows for complicated queries to complete in seconds rather than minutes.

## Data reproducibility

The codebase for scEiaD is available at github.com/davemcg/scEiaD. There are three relevent Snakemake pipelines which were used to create scEiaD: SnakeQUANT, which was used to quantify the gene / cell expression from the raw fastq files, SnakePOP, which runs the multi-parameter integration methods, and SnakeSCEIAD, which takes the optimal parameters from SnakePOP to build the scEiaD sqlite database for the app.

@@

## Updates to the scEiaD backend
Our first version of the single cell Eye is a Disk (scEiaD) database contained `r meta_filter_v0$study_accession %>% unique() %>% length()` studies, three species, `r nrow(meta_filter_v0)` cells, and `r meta_filter_v0$CellType_predict %>% unique() %>% length()` curated cell types. We updated the scEiaD resource for the PLAE v0.90 web app in six important ways. First, we added two chicken retina (*Gallus gallus*) studies. Second we added an ocular outflow tract dataset and two more cornea datasets to enhance coverage across the eye. Third, we added a human pan-body reference scRNA dataset to allow for non-ocular comparisons. Fourth, we used an *in silico* method to remove background gene contamination as we noticed persistent Rhodopsin expression in many non-photoreceptors cells (sup fig XX). Fifth, we increased the minimum number of detected unique transcripts per cell required for a cell to be retained from 200 to 300, which reduces the number of cells that pass QC but improves the quality. Sixth, we carried over and harmonized several common cell type labels from the Tabula Muris dataset to more easily provide non eye cell type comparisons.

After the dataset and quality control updates the scEiaD v2022-03-21 dataset now contains `r meta_filter$study_accession %>% unique() %>% length()` studies, four species, `r nrow(meta_filter)` cells, and `r meta_filter$CellType_predict %>% unique() %>% length()` curated cell types (Fig 1).

@@

While PLAE and Spectacle (https://singlecell-eye.org) both contain large numbers of ocular-related scRNA datasets, they are structured very differently. PLAE is built around the scEiaD database, which runs an extensive multi-stage pipeline to build and identify a high performing scVI-based eye model that is used to integrate all datasets together. This allows for gene queries to be constructed and analyzed across studies. In contrast, Spectacle, as of March 2022, is a compilation of study-level datasets and thus you cannot run queries across studies in Spectacle.

There are a few other differences in between PLAE and Spectacle. First, there are no non-ocular datasets in Spectacle. Second, the cell types labels have not been harmonized. Third, as this resource uses Seurat objects, the app takes many seconds to load an individual dataset and many more to display the data.

@@

There are two interwoven reasons that PLAE is responsive, despite the huge amount of data present. First, the data structure of scEiaD is a sqlite database with pre-calculated tests stored in tables. A sqlite database allows for huge amounts of data to be stored for quick retrieval with minimal loading time. The major downside is that this requires a large amount of disk storage, which makes local (on computer) usage onerous. Second, the visualizations in PLAE (e.g. the UMAP view) pull from the custom database directly and use highly optimized plotting packages (scattermore) to reduce drawing time. Starting from the loading page of Spectacle, it takes approximately six seconds to plot the cell type metadata for a few thousand cells across across a single study. In the same amount of time, PLAE can plot the cell type labels for over a million cells.

## Rich web-available visualizations

PLAE has several ways to visualize the data: one being the 2D UMAP view which, side-by-side, shows gene - cell expression and metadata - cell values. So thus a user can, on the left, see Crx expression across over a million cells, and on the right, see cell type labels for those same million cells. The 2D UMAP view also has a zoom function, so a user can focus on a subset of cells. Clicking on the plot will show metadata for the five nearest cells to the click, which is useful to seeing what kind of cells are present in a small area.

For users interested in seeing how expression of a gene relates to study or other covariates, the "Exp Plot" view displays expression by cell type, predicted cell type, or cluster. The values can then be split and colored by multiple covariates like study or organism. This allows a user to see whether a gene pattern is consistent across studies or species. We also offer a "Dot Plot" view which can display a large number of genes in a space efficient manner. This allows a users to see whether a set of genes has related (or not) expression patterns across multiple cell types.

While scEiaD does not have spatial scRNA seq datasets as of 2022, we do leverage existing knowledge of the structure of the retina to show *in silico* visualization of genes by cell type / layer in a cartoon cross section of the retina.

## Differential testing identifies many genes consistently enriched in various ocular cell types

## Ocular specific gene signatures

## Fovea vs peripheral photoreceptor validation

Sanes proposed X genes (https://www.nature.com/articles/s41598-020-66092-9 --> Figure 9)

PRPH2 and RS1 for cones
NR2E3, SLC24A1, PDE6H, GNAT1 for rods
Also did RGC, Microglia, Astrocytes, Muller Glia

## Design of scEiaD allows for projection of outside data

scEiaD is built using scvitool's scVI / scANVI batch correction tool. As part of the process a computational model of the reference datasets are built. This model is only a few mb in size and can be used on a modest computer to project private scRNA datasets onto it. This allows an ocular researcher several functions. First, the ability to check of quality control of an existing dataset by looking at the distribution of their cells overlapped with our labelled cell types. Second, in perturbation experiments, to identify unusual distributions of cells. Third, to leverage our extensive pre-labelled cell information to transfer cell type labels onto private data. 
# Conclusion
